---
title: "The most damaging and economically costly weather events in the United States (1950-2011)"
author: "Nicolás Ratto"
date: "26-08-2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synopsis

This document explores the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. This will try to answer the questions of which types of events are most harmful to population health and of which types of events have the greatest economic consequences. This analysis will include a time perspective to understand if the most harmful and with the greatest economic consequences events have changed in the period 1950-2011 that the database addresses.

# Data processing


```{r warning=FALSE, message = FALSE}

## Install and load packages

packages = c("tidyverse", "lubridate","ggplot2","data.table","ggpubr")

## Now load or install & load all
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)



# Load data
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
              "StormData.csv.bz2", method = "libcurl")

data<-read.csv("StormData.csv.bz2", header = TRUE, sep = ",")

head(data)

```



```{r warning=FALSE, message = FALSE}
## Change format variables
data$BGN_DATE<-mdy_hms(as.character(data$BGN_DATE))
data$BGN_YEAR<-year(data$BGN_DATE)
data$END_DATE<-mdy_hms(as.character(data$END_DATE))
```


# Results

```{r warning=FALSE, message = FALSE}
#* Does the analysis address the question of which types of events are most harmful to population health?
data %>% group_by(EVTYPE) %>% summarise(FATALITIES=sum(FATALITIES),
                                        INJURIES=sum(INJURIES),
                                        PROPDMG=sum(PROPDMG),
                                        CROPDMG=sum(CROPDMG)) %>% arrange(-FATALITIES,
                                                                          -INJURIES,
                                                                          -PROPDMG,
                                                                          -CROPDMG)
```




```{r warning=FALSE, message = FALSE}

## Agrupate base by event type
a<-data %>% group_by(EVTYPE) %>% summarise(FATALITIES=sum(FATALITIES),
                                        INJURIES=sum(INJURIES),
                                        PROPDMG=sum(PROPDMG),
                                        CROPDMG=sum(CROPDMG),
                                        MAG=sum(MAG)) %>% arrange(-FATALITIES)

## Graphs

g1<-a %>% filter(FATALITIES>200) %>% ggplot(aes(x=reorder(EVTYPE,FATALITIES), y=FATALITIES)) +
  geom_bar(stat="identity", fill="green", colour="black", width=0.8) + 
  scale_y_continuous(labels=function(x) format(x, big.mark = ",", scientific = FALSE)) +
  theme_bw() +  coord_flip() + 
  geom_text(aes(label = format(FATALITIES, big.mark = ",", scientific = FALSE)), hjust=-0.1, colour = "black", size=3.0) +
  labs(title="Fatalities",
       x="EVTYPE", 
       y = "FATALITIES")

g2<-a %>% filter(INJURIES>1000) %>% ggplot(aes(x=reorder(EVTYPE,INJURIES), y=INJURIES)) +
  geom_bar(stat="identity", fill="blue", colour="black", width=0.8) + 
  scale_y_continuous(labels=function(x) format(x, big.mark = ",", scientific = FALSE)) +
  theme_bw() +  coord_flip() + 
  geom_text(aes(label = format(INJURIES, big.mark = ",", scientific = FALSE)), hjust=-0.1, colour = "black", size=3.0) +
  labs(title="Injuries",
       x="EVTYPE", 
       y = "INJURIES")

g3<-a %>% filter(PROPDMG>20000) %>% mutate(PROPDMG=PROPDMG/10) %>% 
  ggplot(aes(x=reorder(EVTYPE,PROPDMG), y=round(PROPDMG))) +
  geom_bar(stat="identity", fill="purple", colour="black", width=0.8) + 
  scale_y_continuous(labels=function(x) format(x, big.mark = ",", scientific = FALSE)) +
  theme_bw() +  coord_flip() + 
  geom_text(aes(label = format(round(PROPDMG), big.mark = ",", scientific = FALSE)), hjust=-0.1, colour = "black", size=3.0) +
  labs(title="Property damage (divided into 10)",
       x="EVTYPE", 
       y = "PROPDMG")

g4<-a %>% filter(CROPDMG>5000) %>% mutate(CROPDMG=CROPDMG/10) %>% 
  ggplot(aes(x=reorder(EVTYPE,CROPDMG), y=round(CROPDMG))) +
  geom_bar(stat="identity", fill="red", colour="black", width=0.8) + 
  scale_y_continuous(labels=function(x) format(x, big.mark = ",", scientific = FALSE)) +
  theme_bw() +  coord_flip() + 
  geom_text(aes(label = format(round(CROPDMG), big.mark = ",", scientific = FALSE)), hjust=-0.1, 
            colour = "black", size=3.0) +
  labs(title="Crop damage (divided into 10)",
       x="EVTYPE", 
       y = "CROPDMG")

g5<-ggarrange(g1,g2,g3,g4,ncol = 2,nrow = 2)
annotate_figure(g5, top=text_grob("The most damaging and economically costly weather events in the United States (1950-2011)"),
                bottom = text_grob("National Climatic Data Center Storm Events",
                                       hjust = 1, x = 1, size = 10))


```






```{r warning=FALSE, message = FALSE}
data %>% filter(EVTYPE %in% c("TORNADO","EXCESSIVE HEAT","FLASH FLOOD",
                              "FLOOD","TSTM WIND",
                              "HAIL")) %>% 
  group_by(EVTYPE,BGN_YEAR) %>% summarise(FATALITIES=sum(FATALITIES),
                                          INJURIES=sum(INJURIES),
                                          PROPDMG=(sum(PROPDMG)/10),
                                          CROPDMG=sum(CROPDMG)/10) %>% 
  gather(key = "type", value = "value", -EVTYPE,-BGN_YEAR) %>% 
  ggplot(aes( x = BGN_YEAR, y =value, fill = type, color = type)) + facet_wrap(~EVTYPE) + 
  geom_line() + theme_bw() + 
  labs(title="The most damaging and economically costly weather events in the United States in time (1950-2011)",
       x="Year",
       y = "Fatalities, injures, property damage and crop damage",
       caption = "National Climatic Data Center Storm Events") + 
  theme(legend.position = "bottom")
```





