---
title: "Storms and other severe weather events"
author: "Nicolás Ratto"
date: "25-08-2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synopsis


```{r warning=FALSE, message = FALSE}

## Install and load packages

packages = c("tidyverse", "lubridate","ggplot2")

## Now load or install & load all
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)



# Load data
act<-read.csv("Input/data/activity.csv",header = TRUE,sep=",")


```


# Data processing


# Results

# Review criteria

* Has either a (1) valid RPubs URL pointing to a data analysis document for this assignment been submitted; or (2) a complete PDF file presenting the data analysis been uploaded?
* Is the document written in English?
* Does the analysis include description and justification for any data transformations?
* Does the document have a title that briefly summarizes the data analysis?
* Does the document have a synopsis that describes and summarizes the data analysis in less than 10 sentences?
* Is there a section titled "Data Processing" that describes how the data were loaded into R and processed for analysis?
* Is there a section titled "Results" where the main results are presented?
* Is there at least one figure in the document that contains a plot?
* Are there at most 3 figures in this document?
* Does the analysis start from the raw data file (i.e. the original .csv.bz2 file)?
* Does the analysis address the question of which types of events are most harmful to population health?
* Does the analysis address the question of which types of events have the greatest economic consequences?
* Do all the results of the analysis (i.e. figures, tables, numerical summaries) appear to be reproducible?
* Do the figure(s) have descriptive captions (i.e. there is a description near the figure of what is happening in the figure)?
* As far as you can determine, does it appear that the work submitted for this project is the work of the student who submitted it



